### 2019-06-21 Writing/response_to_reviewers.tex

\item{\textcolor{blue}{I have some concerns about the approach, however. I was somewhat confused by the CD metric. When quantifying n, how is it decided how many taxa have "comparable" characters?
If you simulated the character for all taxa, do they not all have this character (i.e. n always equals the dataset size)?
But more importantly, will two characters not be more similar if they share the same number of character states, independent of if they are telling you the same evolutionary history?
It seems to me that whether or not two characters share a state space is not the relevant axis along which to measure correlation.}}
\label{gower}

Using a Gower based distance, we only compared the comparable taxa between characters.
For example, for two characters \texttt{x = {0, 1, 2, 3}} and \texttt{y = {0, ?, ?, ?}}, the Character Difference metric will be effectively calculated using only the first taxon (\texttt{x = {0}} and \texttt{y = {0}}) and would result in a Character Difference of 0.
As the reviewer points out, characters with less taxa in common will thus be less likely to be different.
This is consistent with what the CD metric actually captures in terms of phylogenetic signal (the number of share splits).
In the example above, the both characters imply at least one split (between the first taxon and the others).
In other words, in a phylogenetic software, both characters will support the same split.

As the reviewer also correctly points out, we simulated our data with no missing data so this does not affect our results.
The precision in the manuscript was just describing the properties of the new CD metric.

We've clarified how missing data can be potentially treated in the methods section:
\textit{For example, for two characters X and Y for four taxa, if character \texttt{X = \{0,0,?,1\}} for the first, second, third and fourth taxa respectively and character \texttt{Y = \{0,0,1,1\}}, only the first, second and fourth taxa are compared for this character.
Note that in these simulations, no missing data was included so the characters are compared for equal number of taxa (see below).} lines 212-2178.

@@

We've added the proportion of replaced (i.e. duplicated) characters in each matrices for each ``minimised'', ``maximised'' and ``randomised'' scenarios along with the changes in overall Character Differences in the supplementary material 3, figures 4, 5 and 6.

The maximising/minimising character differences was done \textit{a priori} without any targeted number of characters to replace to make the analysis less idiosynchratic.
In other words, we decided to modify the randomly simulated matrices to globally increase/decrease character differences which is expected to decrease anyway in bigger matrices.
When using a small number of character states (i.e. two or three), the matrices with few taxa and many characters are expected to have an overall low character difference because of the smaller combinatoric possibilities for each characters.

We've added mentions to the proportion of replaced characters analysis in the supplementary and added a paragraph explaining this caveat in the discussion:

\textit{Furthermore, it is important to note that on average, fewer characters were duplicated (replaced) in the ``minimised'' matrices compared to both the ``maximised'' and ``randomised'' matrices (supplementary material 3, Figs 4, 5 and 6).
However, this change in the amount of replaced characters did not changed the way average character differences changed across the three transformed matrices: the character differences were always consistently increased (``maximised''), decreased (``minimised'') and maintained at a normal level (``randomised'') for each replicate (supplementary material 3, Figs 1, 2 and 3).
This suggest that character difference can be decreased with a smaller amount of characters replacements than increasing it.} lines 459-469.

@@

\begin{enumerate}

\item{\textcolor{blue}{"Phylogenetic analysis algorithms require the assumption of character independence - a condition generally acknowledged to be violated by morphological data."\\
I don't love this sentence because it emphasizes morphology as distinct, when we recognize non-independence as a problem in basically all data types (see Huelsenbeck and Nielsen, 1996, Effect of Nonindependent Substitution on Phylogenetic Accuracy and the references therein). I do think you should point to a citation here; Davalos et al. 2014 (Integrating Incomplete Fossils by Isolating Conflicting Signal in Saturated and Non-Independent Morphological Characters) is a good one.}}
\label{molecular}

We slightly rephrased the sentence in the abstract (See reviewer 3 minor revision \ref{abstract}) and expanded on the distinction between both character types this in the introduction following the reviewer pertinent biobliography suggestion:

\textit{This independence is assumed for both molecular and morphological characters (Huelsenbeck and Nielsen, 1999; D\'{a}valos et al., 2014, Zou and Zhang, 2016).
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the number of states analysed (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but may have an arbitrary number of states (up to that supported by phylogenetic software; Guillerme and Cooper, 2016).
Second, correlations have been proposed to be stronger in morphological data (i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread-out'' correlations in molecular data (i.e. ``diffused'' correlations - though some regions of the genome may be thought as more correlated than others; Huelsenbeck and Nielsen, 1999).
Furthermore, because of this ``spread'' of correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function cf. changes in nucleotide states ; see below).} lines 42-57.

@@

We added the suggested citation (line 74).

\item{\textcolor{blue}{Coding dependence:\\
This section could use a figure illustrating these types of dependence, and what they mean. For example: "For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character." is a fairly confusing sentence. I took this to mean that if you are coding, for example, tail characters, you might atomize the tail into a suite of characters, all of which are absent in organisms with no tail. But I'm not sure I've understood your meaning correctly.}}

We do not feel that adding a figure for illustrating the matrix will improve clarity of an already long and thorough description.
However, we've trying to clarify the description with an example as follows:

\textit{For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.} lines 94-95 and \textit{In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.} lines 101-103.

@@

\item{\textcolor{blue}{"Correlation between characters detected by the software:"\\
I really struggled with this section. It seems like the problem here is still a coding problem - that the authors of these four characters have chosen four highly dependent characters to build the matrix. Even scoring by hand under parsimony would have problems here. I think what you need to decide here is what you want to communicate in this paragraph. Is it that software is not currently equipped to handle this issue?}}

We've added a paragraph highlighting how the problem indeed comes from the fact that current software are not equipped to handle this issue:

\textit{In fact, for practical reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} that seem to be more likely to reflect the split by separating artiodactyles from the cetaceans.
In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history (i.e. the same split) but for the wrong reasons (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleen, the same split would still be estimated by the software).
One solution would be to a) weight some characters (e.g. giving \textbf{C3} and \textbf{C4} a lower weight) and b) make some characters dependent on each other (e.g. making \textbf{C3} and \textbf{C4} dependent on \textbf{C1}).
However, this can also be a bad practice since it can easily lead to the resulting tree reflecting the researcher's prior expectations rather than the true evolutionary history.} lines 143-156.

@@

We've clarified this sentence (see Reviewer 1, Major suggestion \ref{gower}).

\item{\textcolor{blue}{"This procedure was used to treat all the characters are unordered with no assumption on the meaning of the character state"\\
Why is this important? The text doesn't mention testing anything related to orderedness, or specifying anything related to orderedness in MrBayes.}}

We removed the mention of orderdness of the characters since, indeed, it might be more confusing than helpful in this context (we didn't not simulate any ordered characters and the CD metric does not consider characters as ordered - see supplementary material 1). We've changed the sentence to the following:

\textit{This procedure allows comparison of characters regardless of the significance given to their tokens (following the \textit{xyz} notation in Felsenstein, 2004; as used in D\'{a}valos et al., 2014).} lines 220-222.

@@

\item{\textcolor{blue}{"The morphological HKY model"\\
Is not real. This is just the HKY model. Goloboff (2017) would dispute that this model does not favor a Bayesian method, since it is still a Markovian model. That O'Reilly et al. used this model to generate binary data does not mean one has to - you could translate each of the four nucleotides to its own discrete character. This would probably get even further from the Mk model assumptions than collapsing the multistate character produced via the HKY model to a binary character.}}

We've enquoted the ``morphological HKY model'' and specified that it is still a Markovian model:
\textit{Note that both models (``morphological HKY'' and Mk) are both based on a Markovian model and only differ in the number of states generated and transformed.
However, the ``morphological HKY model'' is not the one} directly \textit{estimated in the phylogenetic software used in this protocol (MrBayes, see below).} lines 272-276.

@@

%TG: maybe add a bit more?

\item{\textcolor{blue}{"Recently however, Wright et al. (2016) have shown that an equal rate transition is still the most present in empirical data."\\
Of the models tested - it's plausible there are other explainers of the data. I would change this to "have shown that equal transition rate is a reasonable assumption for many empirical datasets"}}

We've changed the sentence as suggest by the reviewer (lines 262-263).

@@

We meant that modifying phylogenetic matrices after the phylogenetic inference is a way to test the effect of the sources of correlation but is probably not a good practice in systematics.
Although re-weighting characters based on consistency or retention indices can be used to objectively improve the topology, changing the matrix's characters \textit{a posteriori}, it can be seen as dishonest to remove characters because they ``don't give the right signal''). We've removed the end of this sentence however to avoid any further confusion.

\item{\textcolor{blue}{"However, we do not compare the ``maximised'', ``minimised'' and ``randomised'' to the ``true'' tree but rather to the ``normal'' tree. "\\
I understand why you did this, but I think you should also compare normal to true. That will give us some idea of the scope of how close we can get to true, under these conditions, and will help contextualize the results.}}

We've added the comparisons between the ``normal'' and ``randomised'' trees to the ``true'' tree in the supplementary material 3, figures 9 to 11.
We've added a reference to these figures in the main text:

\textit{However, we did report the comparison between the ``true'' trees and both the ``normal'' and ``randomised'' in the supplementary materials 3, Figs 12, 13 and 14.} lines 347-3478.

@@

Therefore, if cumulating all the seeds used in the generation of the matrices (trees + characters) and the modifications of the matrices is likely to be no more than $6.4e^{7}$.
Therefore we do not believe that the random number generation was a proper since we sampled two orders of magnitude less random numbers than the total ones available.

\item{\textcolor{blue}{How did you ``randomly'' modify your matrices? Did you sample under some distribution? If so, which one?}}

Every replaced characters were randomly selected randomly from the total sample of characters (i.e. randomly sampling from the uniform integers distribution (1, $c$); where $c$ is the number of remaining characters). We have now specified this in the text:

\textit{Each replaced characters (i.e. the characters with a weight of +$1$) were randomly selected by sampling them randomly from the list of the remaining characters.} lines 298-300.

@@

There is actually a vigorous debate surrounding how to generate biologically meaningful morphological characters \citep[][]{OReilly20160081,goloboff2017weighted,puttick2017uncertain,GoloboffEmpirical,OReillyEmpirical}.
This ongoing debate, however, focuses mainly on inference methods where data simulation is a tangential aspect of it.
Therefore, we decided to base our simulation protocol on the empirical observations showing that half of 206 analysed datasets supported a M$k$-like model \citep[][Fig. 3]{Wright01072016} and the other half being a combination of models not explicitly described here but that are believed to be covered with the HKY-binarised model \citep{puttick2017uncertain,OReillyEmpirical}.
In our simulations, the idea was to cover a broad range of morphological characters (with variable state numbers, evolution and transition rates) in order to capture a realistic image of the ``discrete morphological character space''.

\item{\textcolor{blue}{Similarly, why did you choose 0.85 for the Mk model? Does it represent some important set of character sequences? How does it represent the space in general?}}

This 0.85 proportion of binary characters and 0.15 of three state ones was extracted from the 100 empirical matrices analysed in \cite{Guillerme2016146} (Supplementary appendix 1. Appendix A: Tree inference) and \cite{ZouConvergence}.
We've specified this in the text:

\textit{We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters (from empirical observations in: Guillerme and Cooper, 2016b; Zou and Zhang, 2016).} lines 257-259.

@@

\begin{enumerate}

\item{\textcolor{blue}{I feel that the manuscript is certainly worth publishing in Systematic Biology, although I worry that the simulation procedure used by the authors limits their ability to thoroughly test of the effect of between character correlation on the accuracy of estimating phylogenetic trees. The authors note this themselves in the discussion, but I feel more should be made of this point. If the simulations utilised an explicit distribution of character covariance then I think that interpretation of the results would have been more straightforward.}}
\label{stramlined}

As noted in the introduction, the aim of this manuscript was mainly to see does different levels of correlation between discrete morphological characters affect tree topology.
We also review different sources of correlation for characters and discuss how the multidimensional complexity of a phenotype (and the correlations therein) is not reflected in two dimensional character matrices.
We do agree with the reviewer that a thorough analysis of the variance/covariance structure of simulated ``discretised'' characters (see reviewer 3, major suggestion \ref{hiden}) would have broad more insight into the nature of character correlation on phylogenetics inference in general (regardless of the type of character and correlation) but due to the length required to run the current ``simplistic'' simulation, we feel that this would be out of the scope of this manuscript.

We have now clarified in the manuscript what our objectives were by streamlining the last paragraph of the introduction (lines 159-170).
We have also explicitly added in the supplementary material 3, figs 1, 2 and 3 how the character difference changes between all matrices compared to the ``normal'' matrix.

@@

\item{\textcolor{blue}{Another worry I have is that the implied covariance at the tips does not necessarily reflect covariation over the full phylogeny. It is conceivable that two characters can appear to have near perfect correlation at the tips, but arrived at these states through completely independent pathways, particularly when simulating with a continuous time Markov chain where unobserved transitions to a new state and back to the original state are possible.}}
\label{hiden}

We agree with this reviewer's comment which can be especially thought to be the case for discrete characters with a fast rate of evolution.
Such characters will likely swap between states along the branches following a hidden Markov chain.

However, we feel that modelling this process might not be useful in this specific study:
First, in empirical data two characters can be strongly correlated for biological and evolutionary reasons (e.g. the absence of an astragalus and the presence of baleen in table 1) leading to a strong correlation of both characters at the tips (or even nodes) but does not mean that both character's hidden Markov chain are correlated.
Second, by comparing characters at any time point rather than simply at the tips (which can be problematic since our trees were not all ultrametric) we would probably not be able to compare the effect of the ``true'' characters' correlation since this information will not be used by the phylogenetic software (MrBayes and PAUP*) that will only use tip values.

We've added this remark in the caveats section:
\textit{It is also worth noting that character similarity at the tips does not necessarily reflects identical evolutionary history.
This may be specifically true when simulating characters based on the M$k$ model where the resulting tip states does not necessarily reflects variation of the state along the branches (e.g. Revell, 2014). However, it is worth noting that in empirical studies, such state changes along the branches are nearly always empirically unknown (even though they can be estimated by Maximum Likelihood).} lines 563-568.

@@

We hope to expand this study in the future by create different groups of characters sharing the same overall correlation by generating ``bunches'' of characters based on different rates (to simulate rate homogeneity) and different trees (branch length and topology - to simulate variation in ``gene'' trees).

\item{\textcolor{blue}{A discussion of the ``effective'' number of characters might also be warranted. When maximising character covariance this likely requires the resampling of the same handful of characters multiple times, each of which provides the exact same phylogenetic signal. The redundancy of these characters seems likely to influence the accuracy of estimates. The choice of characters that are sampled by the bootstrap procedure will therefore constrain the accuracy of topology estimates. If the resampled matrix consists of a handful of characters with high consistency with the true tree, sampled many times over, then the tree estimate will be good (for at least part of the tree) and covariance will be high. Similarly, if the handful of regularly resampled characters are mainly inconsistent (with the true tree) then tree estimates will be poor but the implied covariance will also be high. I think this is evident from the reduced performance in larger matrices, where there will be a significant reduction in phylogenetic signal when attempting to maximise covariance.}}

We've added the following part to the discussion:

\textit{Overall, it is expected that minimising character difference (increasing character correlation) is likely to have a greater effect on recovering the ``correct'' topology since most characters will provide the same phylogenetic signal.
If the ``minimised'' matrix consists of a handful of characters with high consistency with the ``best'' tree, then the tree estimate will be good (for at least part of the tree).
Similarly, the characters in the ``minimised'' matrix are mainly inconsistent with the ``best'' tree then the estimates will be poor.} lines 475-481.

@@

\item{\textcolor{blue}{3:25 – I think that resurgence should be taken out of quotation marks, as no reference is supplied for the use of the phrase}}

We've removed the quotes.

\item{\textcolor{blue}{3:29 – In what ways are morphological data inferior? I think this needs to be qualified.}}

We've added the following justification (and toned down the sentence):
\textit{Although morphological character data are sometimes considered inferior to molecular sequence data due to the difficulties of collected them and their sometimes subjectivity, they are often the only source of phylogenetic data for extinct species.} lines 28-30.

### 2019-06-21 Writing/draft.tex
Finalised version done up to the discussion

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Characters correlation

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf Influence of different modes of morphological character correlation on phylogenetic tree inference}
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,2,*}$, and Martin D. Brazeau$^{2,3}$}\\
\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\
$^2$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\
$^3$Department of Earth Sciences, Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}

@@

\newpage
\begin{abstract}
Phylogenetic analysis algorithms require the assumption of character independence - a condition generally acknowledged to be violated by morphological data.
Correlation between characters can originate from intra-organismal features, shared phylogenetic history or forced by particular character-state coding schemes. 
Although the two first sources can be investigated by biologists \textit{a posteriori} and the third one can be avoided \textit{a priori} with good practices, phylogenetic software do not distinguish between any of them.

In this study, we propose a new metric of raw character difference as a proxy for character correlation.
Using thorough simulations, we test the effect of increasing or decreasing character differences on tree topology.
Overall, we found an expected positive effect of reducing character correlations on recovering the correct topology.
However, this effect is less important for matrices with a small number of taxa (25 in our simulations) where reducing character correlation is not more effective than randomly drawing characters.
Furthermore, in bigger matrices (350 characters), there is a strong effect of the inference method with Bayesian trees being consistently less affected by character correlation than maximum parsimony trees.

These results suggest that ignoring the problem of character correlation or independence can often impact topology in phylogenetic analysis.
However, encouragingly, they also suggest that, unless correlation is actively maximised or minimised, probabilistic methods can easily accommodate for a random correlation between characters.

@@

The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Morphological character data are often considered inferior to molecular sequence data, but are often the only source of phylogenetic data for extinct species. 
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to ``misbehave''.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
Especially for discrete morphological data, this assumption of independence is probably violated frequently due to the very nature of phylogenetic data: correlations are expected to occur (to some degree) when characters are \hl{dependent} on each other.
Before discussing character correlation further, it is important to understand that it may manifest itself in at least three distinct ways:

\begin{itemize}
    \item \textbf{Intra-organismal dependence:} this is the result of an intrinsic biological link between two characters through development, pleiotropy, and/or biological function.
    For example the lower and upper molar characters in mammals generally occlude one another. 
    Therefore, one character describing a feature of a lower molar will be expected to be complemented by the surface of the occluding upper molar.
    Characters of the occlusal surface of two opposing molars will be expected to directly covary.
    Pleiotropy also results in covariation between different aspects of phenotype. 
    From a phylogenetic perspective, it can be especially pernicious because the relationship between the traits in question may have no obvious link from a morphological or functional comparison alone. 
    Intra-organismal links can be the targets of comparative developmental biology \citep{goswami2010,kelly2010reduced,stoessel2013morphological,goswami2014macroevolutionary} or functional investigations.

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving due to selection, likely related to functional links between two traits that help serve an overall lifestyle trait. 
    Unlike the case of intra-organismal dependence, there need not be an intrinsic constraint that causes these traits to covary.
    For example, in vertebrates, axial elongation can be correlated to limb reduction with snake-like bodies evolving multiple times in numerous tetrapod lineages.
    This is thought to correspond to adaptations for fossoriality or aquatic lifestyles.
    Such covariances are generally studied in the context of a given phylogeny, often one derived from molecular data with the morphological traits of interest mapped on it.
    Many methods have been developed to study these correlations, especially since they can provide us with a lot of information on how specific groups acquired specific characteristics \citep{Lande1983,Maddison1990,Pagel1994,Pagel2006,Grabowski2016}.
    However, these methods do not give us a means to objectively control correlations that might adversely affect phylogenetic inference.

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    Coding dependence manifests itself in several ways, particularly in coding redundant information.
    For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character.
    This occurs when a number of multistate characters include two variable feature states (e.g. large, small; red, blue etc.) in conjunction with absence.
    It is worth noting, however, that these correlations could also be due to the nature of the available data, especially in palaeontology.
    For example, when only one fragmentary molar is available to describe a specimen, researchers have to ``extract'' as much phylogenetic information from the available data as possible, potentially inducing correlations.
    This coding dependency is linked to hierarchical dependency between characters \citep{wilkinson1995coping,BrazeauNA}.
    Finally this can also be due to a bias in the amount of characters available.
    For example, in skulls, because of their complexity, there is a high likelihood of inducing correlation (by effectively reducing structural complexity to discrete characters).
\end{itemize}

\noindent Of course, the three sources of dependence have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Logical dependence, however, is easily distinguished prior to phylogenetic inference, while the two other ones (intra-organismal and evolutionary) are much harder.
However, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{DeLaet2015,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (and possibly beyond the limits of available technology).
Even after all of the effort is expended, the results might then only be known for a single (model) species.
Evolutionary dependence itself requires the resolution of a phylogenetic tree, and is best determined by independent character sets. 
This is frequently accomplished by mapping morphological traits on molecular phylogenetic trees.

These sources of dependence between characters are well studied in biology.
Biological and evolutionary dependences are inherent parts to Evo-Devo and macroevolutionary studies and best practices to avoid coding-induced dependences are commonly known and applied.
However, eventually, all these characters, whether they are independent or not are analysed through phylogenetic inferences software that are blind to these distinctions.
If fact, what the software are confronted with is a two dimensional matrix problem that renders the morphological subtleties described opaque.
This introduces a new, less studied, source of character dependence:

\noindent \textbf{Correlation between characters detected by the software}: this is the result how software actually interprets the differences between characters.
The vast majority of phylogenetic software ignores both the character's definition and the different states signification (simply treating them as different or similar tokens).
Therefore a great number of characters and - traditionally - a few number of tokens can easily lead to dependence between characters.
For example, if we consider the following matrix containing four cetartiodactyls - say a pig (e.g. \textit{Sus}), a deer (\textit{Cervus}), a hippo (\textit{Hippopotamus}) and a whale (\textit{Balaenoptera}) - and four binary characters - say (\textbf{C1}: presence (1) or absence (0) of an astragalus; \textbf{C2}: presence (0) or absence (1) of baleen; \textbf{C3}: presence (0) or absence (1) of a left astragalus with a double pulley; \textbf{C4}: presence (0) or absence (1) of a right astragalus with a double pulley:

@@

    \item \textbf{Modifying matrices}: we changed the ``normal'' matrices by modifying the characters in order to maximise or minimise characters differences (hereafter called respectively ``maximised'' and ``minimised'' matrices) by removing respectively the least different or most different characters and replacing them randomly by the remaining characters. Our protocol for measuring character difference is detailed below.

@@

\item \textbf{Inferring topologies}: we inferred the topologies from the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices using both maximum parsimony and Bayesian inference.
Hereafter, the resultant topologies are called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' trees).
This step resulted in 72 topologies.
    \item \textbf{Comparing topologies}: finally, we compared the ``normal'' to the ``maximised'', ``minimised'' and ``randomised'' trees to measure the effect of character correlation on topology.
\end{enumerate}
Each step was replicated 35 times and are described below in more detail, along with our proposed definition for measuring the difference between characters.

>>>>

\begin{document}
%Running head
\begin{flushright}s
Version dated: \today
\end{flushright}

\bigskip
\noindent RH: Characters correlation
\bigskip
\medskip
\begin{center}
\noindent{\Large \bf Influence of different modes of morphological character correlation on phylogenetic tree inference}
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,2,*}$, Abigail I. Pastore$^{1}$, and Martin D. Brazeau$^{2,3}$}\\
% Abigail I. Pastore$^{1}$
\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\
$^2$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\
$^3$Department of Earth Sciences, Natural History Museum, Cromwell Road, London, SW75BD, United Kingdom.\\}

@@

% CITE: https://academic.oup.com/sysbio/article/68/2/267/5145070#130574594

% Mention Hamming distance rather than Gower?

\newpage
\begin{abstract}
Phylogenetic analysis algorithms assume character independence - a condition generally acknowledged to be violated by morphological data.
Correlation between characters can originate from intra-organismal features, shared phylogenetic history or forced by particular character-state coding schemes. 
Although the two first sources can be investigated by biologists \textit{a posteriori} and the third one can be avoided \textit{a priori} with careful character coding, phylogenetic software do not distinguish between any of them.
In this study, we propose a new metric of raw character difference as a proxy for character correlation.
Using thorough simulations, we test the effect of increasing or decreasing character differences on tree topology.
Overall, we found a negative effect of reducing character correlations on recovering the correct topology.
However, this effect is less important for matrices with a small number of taxa (25 in our simulations) where reducing or increasing character correlation is not more effective in getting the expected topology than randomly drawing characters.
Furthermore, in bigger matrices (350 characters), there is a strong effect of the inference method with Bayesian trees being consistently less affected by character correlation than maximum parsimony trees.
These results suggest that ignoring the problem of character correlation or independence can often impact topology in phylogenetic analysis.
However, encouragingly, they also suggest that, unless correlation is actively maximised or minimised, probabilistic methods can easily accommodate for a random correlation between characters.

@@

The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Although morphological character data are sometimes considered inferior to molecular sequence data due to the difficulties of collecting them and their sometimes subjectivity, they are often the only source of phylogenetic data for extinct species.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to ``misbehave''.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
This independence is assumed for both molecular and morphological characters \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the number of states analysed (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but may have an arbitrary number of states (up to that supported by phylogenetic software; \citealt{Guillerme2016146}).
Second, correlations have been proposed to be stronger in morphological data (i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread-out'' correlations in molecular data \citep[i.e. ``diffused'' correlations - though some regions of the genome may be thought as more correlated than others][]{huelsenbeck1999effect}.
Furthermore, because of the ``spread'' of correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).
Before discussing character correlation further, we distinguish three different modes of forms of character correlation:

\begin{itemize}
    \item \textbf{Intra-organismal dependence:} this is the result of an intrinsic biological link between two characters through development, pleiotropy, and/or biological function.
    For example the lower and upper molar characters in mammals generally occlude one another. 
    Therefore, one character describing a feature of a lower molar will be expected to be complemented by the surface of the occluding upper molar.
    Characters of the occlusal surface of two opposing molars will be expected to directly covary.
    Pleiotropy also results in covariation between different aspects of phenotype and may thus induce correlations. 
    From a phylogenetic perspective, it can be especially pernicious because the relationship between the traits in question may have no obvious link from a morphological or functional comparison alone. 
    Intra-organismal links can be the targets of comparative developmental biology \citep{goswami2010,kelly2010reduced,stoessel2013morphological,goswami2014macroevolutionary} or functional investigations.

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving due to selection, likely related to functional links between two traits that help serve an overall lifestyle trait \citep{ClarkeMiddleton2008}. 
    Unlike the case of intra-organismal dependence, there need not be an intrinsic constraint that causes these traits to covary.
    For example, in vertebrates, axial elongation can be correlated to limb reduction with snake-like bodies evolving multiple times in numerous tetrapod lineages.
    This is thought to correspond to adaptations for fossoriality or aquatic lifestyles.
    Such covariances are generally studied in the context of a given phylogeny, often one derived from molecular data with the morphological traits of interest mapped on it.
    Many methods have been developed to study these correlations, especially since they can provide us with a lot of information on how specific groups acquired specific characteristics \citep{Lande1983,Maddison1990,Pagel1994,Pagel2006,Grabowski2016}.
    However, these methods do not give us a means to objectively control correlations that might adversely affect phylogenetic inference.

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    Coding dependence manifests itself in several ways, particularly in coding redundant information.
    For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character.
    This occurs when a number of multistate characters include two variable feature states (e.g. large, small; red, blue etc.) in conjunction with absence.
    For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.
    It is worth noting, however, that these correlations could also be due to the nature of the available data, especially in palaeontology.
    For example, when only one fragmentary molar is available to describe a specimen, researchers have to ``extract'' as much phylogenetic information from the available data as possible, potentially inducing correlations.
    This coding dependency is linked to hierarchical dependency between characters \citep{wilkinson1995coping,BrazeauNA}.
    In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.
    Finally this can also be due to a bias in the amount of characters available.
    For example, in skulls, because of their complexity, there is a high likelihood of inducing correlation (by effectively reducing structural complexity to discrete characters).
\end{itemize}

\noindent Of course, the three sources of dependence have an interaction: characters describing the left and right lower/upper molars will have induced dependence due to the modularity of the molars, their shared history and the duplicated coding.
Logical dependence, however, is easily distinguished prior to phylogenetic inference, while the two other ones (intra-organismal and evolutionary) are much harder.
However, the development of algorithms and software has not yet caught up with the need to deal with these interdependencies \citep{de2015parsimony,BrazeauNA}.
Intra-organismal dependence requires more detailed, often extremely time-consuming studies (and possibly beyond the limits of available technology).
Even after all of the effort is expended, the results might then only be known for a single (model) species.
Evolutionary dependence itself requires the accurate resolution of a phylogenetic tree, and is best determined by independent character sets. 
This is frequently accomplished by mapping morphological traits on molecular phylogenetic trees.

These sources of dependence between characters are well studied in biology.
Biological and evolutionary dependences are inherent parts to Evo-Devo and macroevolutionary studies and best practices to avoid coding-induced dependences are commonly known and applied.
However, eventually, all these characters, whether they are independent or not are analysed through phylogenetic inference software that are blind to these distinctions.
If fact, what the software are confronted with is a two dimensional matrix problem that renders the morphological subtleties described opaque.
This introduces a new, less studied, source of character dependence:

\noindent \textbf{Characters compatibility}: this is the result of how software actually interprets the differences between characters.
The majority of phylogenetic algorithms cannot really interpret character descriptions and state meanings (i.e. character ontologies), and instead must interpret discrete tokens as signifiers of shared similarity.
Therefore, phylogenetic character matrices that contain a great number of characters but a few number of tokens can easily lead to dependence between characters.
For example, if we consider the following matrix containing four cetartiodactyls - say a pig (e.g. \textit{Sus}), a deer (\textit{Cervus}), a hippo (\textit{Hippopotamus}) and a whale (\textit{Balaenoptera}) - and four binary characters - say (\textbf{C1}: presence (1) or absence (0) of an astragalus; \textbf{C2}: presence (0) or absence (1) of baleen; \textbf{C3}: presence (0) or absence (1) of a left astragalus with a double pulley; \textbf{C4}: presence (0) or absence (1) of a right astragalus with a double pulley.

@@

    \item \textbf{Modifying matrices}: we changed the ``normal'' matrices by modifying the characters in order to maximise or minimise characters differences (hereafter called respectively ``maximised'' and ``minimised'' matrices) by removing respectively the least different or most different characters and replacing them randomly by characters from the remaining set. Our protocol for measuring character difference is detailed below.

@@

\item \textbf{Inferring topologies}: we inferred the topologies from the ``normal'', ``maximised'', ``minimised'' and ``randomised'' matrices using both maximum parsimony and Bayesian inference.
Hereafter, the resultant topologies are called the ``normal'', ``maximised'', ``minimised'' and ``randomised'' trees).
This step resulted in 72 topologies.
    \item \textbf{Comparing topologies}: finally, we compared the ``normal'' to the ``maximised'', ``minimised'' and ``randomised'' trees to measure the effect of character correlation on the recovered topology.
\end{enumerate}
All steps were replicated 20 times and are described below in more detail, along with our proposed definition for measuring the difference between characters.
